Detecting Photo Manipulation using Reflections on Curved Surfaces

### 2014-02-10 paper/intro.tex
results section in progress

% TODO how sensitive are humans to seeing inconsistent reflections? (see obrien12)

Image forensics, the study of finding hidden information in photographs, is an emerging field with diverse applications. It is comprised of a few general subtopics: determining the source of an image, detecting forgery, and detecting steganography (hiding information in a digital image) \cite{rocha2011}. All too often, pictures are used as hard evidence in journalism, scientific research, and criminal cases. Image forgery is nothing new, but as computers have become more capable and tools such as Adobe Photoshop have been developed, it has become much easier to manipulate images in a convincing manner. Therefore, there is a need for forensic techniques.

Existing forensic techniques generally involve finding statistical anomalies, machine learning, or finding geometric inconsistencies. The current techniques that take advantage of reflective objects in an image fall into the latter category.

Most of the reflective surfaces that might be found in real photos are planar. Some common examples are bodies of still water, mirrors, and glass windows. Non-planar reflective surfaces are less common, but can be found in the form of store security mirrors, works of art, and even the human eye.

>>>>

Image forensics, the study of finding hidden information in photographs, is an emerging field with diverse applications. It is comprised of a few general subtopics: determining the source of an image, detecting forgery, and detecting steganography (hiding information in a digital image) \cite{rocha2011}. All too often, pictures are used as hard evidence in journalism, scientific research, and criminal cases. For example, there was one incident in the scientific community in which a group of South Korean scientists published groundbreaking stem cell research results, but it was later revealed that the majority of the photographs included with the results were doctored \cite{rocha2011}. Image forgery itself predates the digital age, but as computers have become more capable and tools such as Adobe Photoshop have been developed, it has become much easier to manipulate images in a convincing manner. Therefore, there is an increasing need for forensic techniques as the technology develops.

Existing forensic techniques generally involve machine learning, finding statistical anomalies, or finding geometric inconsistencies. The current techniques that take advantage of reflective objects in an image fall into the latter category. Interestingly, the human eye is very insensitive to inconsistencies in reflections \cite{farid2010image}. Therefore it would be fairly straightforward for a forger to, for example, convincingly show a person in a scene with a reflective surface who was never actually there. It would perhaps be convincing to a human, but may not hold up to analysis with a geometric technique.

Most of the reflective surfaces that might be found in real photos are planar. Some common examples are bodies of still water, mirrors, and glass windows. Non-planar reflective surfaces, which are the focus of this paper, are less common but can be found in the form of store security mirrors, works of art, and even the human eye.

### 2014-02-10 paper/main.tex
results section in progress

\documentclass{thesis}
%\usepackage{lgrind}
\usepackage{graphicx}
\usepackage{verbatim}

@@

\pagestyle{plain}
\bibliographystyle{apalike}

@@

\chapter{Prior Work}
\section{Reflections on Planar Surfaces}
\begin{figure}
	\centering
    	\includegraphics[width=0.8\textwidth]{planar-vanishing-point-forged}
	\caption{Demonstration of vanishing point inconsistency. The lines corresponding to the man on the left converge to a different vanishing point than those from the rest of the scene, so it is likely that the man and his reflection were edited into the scene. \cite[Fig.~1]{obrien12}}
	\label{planar-vanishing-point-forged}
\end{figure}

@@

\section{Other Related Work}
% TODO Shadows

@@

\section{Properties of Reflective Geometry}
Reflection of light on a surface is directly related to the surface normal at the point of reflection. The incident ray (or ray of light coming towards the surface) must make the same angle with the normal as the reflected ray.  Additionally, the incident ray, reflected ray, and normal must all be coplanar, and the reflected and incident rays must be on opposite sides of the normal.

@@

In the case of a planar mirror, all of these rays would be parallel in three-dimensional space.\nocite{obrien12} An axiom of linear perspective projection is that parallel lines projected onto a 2D image must converge to some vanishing point $V$ \cite[p.~2]{hartley}, so all such lines through a planar mirror intersect at a single vanishing point if the photograph is unaltered.

@@

\begin{figure}
	\centering
    	\includegraphics[width=0.6\textwidth]{colinear-radius-camera-view}
	\caption{A view of Figure~\ref{colinear-radius} from the point of view of the camera $C$.}
	\label{colinear-radius-camera-view}
\end{figure}

@@

From the perspective of the camera, since the projection of the incident ray is colinear with the normal, it must also be colinear with the radius. Therefore, in theory, in a valid picture of a reflective sphere any incident ray must appear to pass through the center of the sphere [Figure~\ref{colinear-radius-camera-view}].

\section{Why not estimate the normal?}
In general, the types of techniques outlined verify that the incident ray is consistent with the normal at the point of reflection. If the projection of the normal could be estimated on any arbitrary image of a quadric surface, then we could easily have a technique to detect inconsistencies in reflections on most curved surfaces because the projection of the normal and the projection of the incident ray should be colinear. However, when perspective projection is taken into account, it is non-trivial to estimate the normal accurately on an arbitrary visible point on the surface, even for a simple surface like a sphere.

In the unrealistic case where an image was produced using orthographic projection, the projection of the normal could be easily estimated on a sphere because the sphere appears exactly the same no matter the angle from which it is viewed, and the entire hemisphere facing the camera is visible. The $x$ and $y$ components of the projection of the normal vector are simply a function of the projection of the position on the sphere. With perspective projection, the visible region of the sphere is slightly smaller than the projection of the hemisphere so the radius can not be determined just from the visible size of the sphere.

@@

\subsection{Distortion from Perspective Projection}
One issue with this naive approach to finding the center is that perspective projection distorts the shape of the sphere, depending on how centered the sphere is in the camera's view. If the sphere is towards the edge of the camera's view, the extent of the visible part of the sphere towards the edge is closer to the camera than the visible extent of the part of the sphere towards the middle of the image. This causes the sphere to look larger towards the edge of the image and therefore the apparent edge of the circle distorts out towards the edge of the image.

% TODO IRL example of this phenomenon
% TODO example with smaller sphere
% TODO example with orthographic projection
% TODO diagram

\chapter{Conclusion}
\section{Results}
% explain the need to plot the center point
% why it's not useful to consider the intersection of points, but rather how close they come to the circle center
% typical results for untampered images- how many pixels away from the center, how scattered the intersections can be
% address: effects of higher resolution, edited images: shifts forward/back, side to side, farther back, composite,
% why the radius test breaks for certain adjustments
%\subsection{Effects of Human Error}
\section{Future Work}

>>>>

\documentclass{thesis}
%\usepackage{lgrind}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{verbatim}

@@

\pagestyle{plain}
\bibliographystyle{apalike}
\renewcommand{\bibname}{References}

@@

\chapter{Prior Work}
\section{Reflections on Planar Surfaces}
\label{sec:planar-reflections}
\begin{figure}
	\centering
    	\includegraphics[width=0.8\textwidth]{planar-vanishing-point-forged}
	\caption{Demonstration of vanishing point inconsistency. The lines corresponding to the man on the left converge to a different vanishing point than those from the rest of the scene, so it is likely that the man and his reflection were edited into the scene. \cite[Fig.~1]{obrien12}}
	\label{planar-vanishing-point-forged}
\end{figure}

@@

\section{Other Geometric Techniques}
Other forensic techniques exist that take a geometric approach to shadow consistency. Instead of determining if there is a consistent vanishing point, or a plausible consistent camera location, these techniques determine if there is a consistent location for a light source. One simple technique for determining shadow consistency, akin to the techniques for reflection consistency, involve drawing rays between points on an object and corresponding points on the shadow. These rays should intersect in a plausible location for the light source (or they should be nearly parallel if the light source is the sun) \cite{farid2010image}. This is dependent on the shadows having easily distinguishable features that can be matched up to the object that is casting the shadow.

\begin{figure}
	\centering
    	\includegraphics[width=0.8\textwidth]{moon-landing-consistent}
	\caption{Based on shadow wedge constraints, this photo of the moon landing has a consistent light source. Original image copyright 1969, NASA. \cite[Fig.~1]{kee2013exposing}}
	\label{moon-landing-consistent}
\end{figure}

Another similar technique improves on the previous one by eliminating the need for distinguishable shadow features. Instead of drawing rays, a point is picked on a shadow and a wedge constraint created that contains all of the object that could have plausibly cast that point of the shadow \cite{kee2013exposing}. If there exists a region that satisfies all the constraints, the image passes this test, as in Figure~\ref{moon-landing-consistent}.

% TODO section describing approach? including assumptions and why they are reasonable, overview of tools

@@

\section{Properties of Reflective Geometry}
% TODO expand to sync up nomenclature
Reflection of light on a surface is directly related to the surface normal at the point of reflection. The incident ray (or ray of light coming towards the surface) must make the same angle with the normal as the reflected ray.  Additionally, the incident ray, reflected ray, and normal must all be coplanar, and the reflected and incident rays must be on opposite sides of the normal.

@@

In the case of a planar mirror, all of these rays would be parallel in three-dimensional space.\nocite{obrien12} An axiom of linear perspective projection is that parallel lines projected onto a 2D image must converge to some vanishing point $V$ \cite[p.~2]{hartley}, so all such lines through a planar mirror intersect at a single vanishing point if the photograph is unaltered. This is the mathematical basis for the technique described in Section~\ref{sec:planar-reflections}.

@@

\begin{figure}
	\centering
    	\includegraphics[width=0.6\textwidth]{colinear-radius-camera-view}
	\caption{A view of Figure~\ref{colinear-radius} from the viewpoint of the camera $C$.}
	\label{colinear-radius-camera-view}
\end{figure}

@@

From the view point of the camera, since the projection of the incident ray is colinear with the normal, it must also be colinear with the radius. Therefore, in theory, in a valid picture of a reflective sphere any incident ray must appear to pass through the center of the sphere [Figure~\ref{colinear-radius-camera-view}].

\section{Why not estimate the normal?}
% TODO explain that camera parameters are needed
In general, the types of techniques outlined verify that the incident ray is consistent with the normal at the point of reflection. If the projection of the normal could be estimated on any arbitrary image of a quadric surface, then we could easily have a technique to detect inconsistencies in reflections on most curved surfaces because the projection of the normal and the projection of the incident ray should be colinear. However, when perspective projection is taken into account, it is non-trivial to estimate the normal accurately on an arbitrary visible point on the surface, even for a simple surface like a sphere.

In the unrealistic case where an image was produced using orthographic projection, the projection of the normal could be easily estimated on a sphere because the sphere appears exactly the same no matter the angle from which it is viewed, and the entire hemisphere facing the camera is visible. The $x$ and $y$ components of the projection of the normal vector are simply a function of the projection of the position on the sphere. With perspective projection, the visible region of the sphere is slightly smaller than the projection of the hemisphere so the radius can not be determined from the visible size of the sphere alone.

@@

\subsection{Distortion from Perspective Projection}
\label{sec:perspective-distortion}
\begin{figure}
	\centering
	\begin{subfigure}[b]{0.4\textwidth}
                \includegraphics[width=\textwidth]{sphere-perspective-zoom}
                \caption{Perspective view of a sphere}
    \end{subfigure}
    ~
   	\begin{subfigure}[b]{0.4\textwidth}
                \includegraphics[width=\textwidth]{sphere-ortho}
                \caption{Orthographic view of a sphere}
    \end{subfigure}
    \caption{Viewing the sphere with perspective projection introduces noticeable error when solving for the circle projected onto the image plane.}
	\label{sphere-perspective-ortho}
\end{figure}

One issue with this naive approach to finding the center is that perspective projection distorts the shape of the sphere, depending on how centered the sphere is in the camera's view. If the sphere is towards the edge of the camera's view, the extent of the visible part of the sphere towards the edge is closer to the camera than the visible extent of the part of the sphere towards the middle of the image. This causes the sphere to look larger towards the edge of the image and therefore the apparent edge of the circle distorts out towards the edge of the image. As a result, the outline projected onto the image is slightly elliptical, as in Figure~\ref{sphere-perspective-ortho}.

% TODO IRL example of this phenomenon?
% TODO example with smaller sphere
% TODO diagram

\chapter{Results}
As detailed in Section~\ref{sec:perspective-distortion}, it as not as simple as finding the outline of a perfect circle to find the projection of the center of the sphere in the image. Therefore, for the purpose of testing the method, test images were generated with a small sphere located at the exact center of the reflective sphere made slightly transparent.

Because there is some error associated with having a human input the points to create the projections of the incident rays, it is assumed that in an unaltered image the rays will probably not cross the exact center of the sphere. For this reason, the measure for authenticity will be how close the rays come to the center of the sphere.

One possible alternative to looking at the minimum distance to the center would be to consider the the point of intersection of all the rays, similar to the vanishing point method in the planar case. However, the spherical case is much more sensitive to error than the planar case. In the planar case the incident rays are parallel, so their vanishing point often lies far outside the image, and so the position of the intersection point is not very sensitive to a difference of a few pixels in the points picked. In the spherical case, however, it is not uncommon for the projections of the incident rays to not intersect cleanly even in genuine images.

\section{Typical Results for Genuine Images}
% typical results for untampered images- how many pixels away from the center, how scattered the intersections can be
% address: effects of higher resolution, edited images: shifts forward/back, side to side, farther back, composite,
% why the radius test breaks for certain adjustments

\chapter{Future Work}
